Project Preparation Report\\Bayesian Inference Using Sequential Monte-Carlo Algorithm for Dynamic System Models

### 2020-07-02 MSc_dissertation.tex
Ready for work

The project aims at the parameter estimation and model comparison. Approximate 
Bayesian computation (ABC) approaches will be used to estimate parameters of mathmatical

@@

\begin{equation} \label{eq:1}
\begin{array}{l}
\frac{\mathrm{d} N}{\mathrm{d} t}=\lambda_N+\kappa_{N\beta}\beta-\mu_NN-\nu_{N\Phi}N\Phi\\\\
\frac{\mathrm{d} \Phi}{\mathrm{d} t}=\lambda_\Phi+\kappa_{\Phi\beta}\beta-\mu_\Phi\Phi\\\\
\frac{\mathrm{d} \beta}{\mathrm{d} t}=\frac{s_{\beta N}N}{1+i_{\beta\Phi}}-\mu_\beta\beta\\\\
\frac{\mathrm{d} \alpha}{\mathrm{d} t}=s_{\alpha\Phi}\Phi-\mu_\alpha\alpha
\end{array}
\end{equation}

A mathematical model would introduce parameters (e.g. $\kappa$, $\mu$ ,etc in (1)), which are to be determined to 
give a specific model using observed data. A traditional method to find the best desired parameter sets 
is to apply an optimisation-based fitting, where the discrepancy of the model’s 
predictions and experimental data is minimised. Alternatively, by adopting Bayesian 
inference, we can focus on the experimental data and try to infer the parameter sets. 
Many approaches have been proposed and used in the area of systems biology\cite{ref:Stumpf}. Compared 
to optimisation-based fitting methods, Bayesian inference approaches can relieve the 
concerns of over-fitting and local optima. It focuses on posterior distribution which 
represents parameters’ ‘uncertainty’ given the observation data. As Bayes rule 
stated, the posterior distribution $p(\theta|D)$ is proportional to product of likelihood and 
prior
\begin{equation} \label{eq:2}
p(\theta|D)\propto l(\theta|D)p(\theta)
\end{equation}
where $\theta$ denotes parameter vector, $l$ is the likelihood function and $D$ is the observed data.

@@

1. pyABC

\quad $\bullet$ Built-in support for multiple types of  parallelisation

\quad $\bullet$ Built-in visualisation

2. astroABC

\quad $\bullet$ Support parallel sampling using MPI or multiprocessing

\quad $\bullet$ Each run can be paused and resumed later

3. ELFI

\quad $\bullet$ Connect to ipyparallel for more parallelisation options

\quad $\bullet$ Using directed acyclic graph as model

4. ABC-SysBio

\quad $\bullet$ Experimental CUDA accelerated parallelisation

5. EasyABC

\quad $\bullet$ Multicore parallelisation

\quad $\bullet$ Offer several different sequential algorithms

6. smfsb

\quad $\bullet$ Provide simple and basic output of parameter estimations.

>>>>

The project aims at the parameter estimation and model comparison. As a kind of 
computational Bayesian inference methods, Approximate 
Bayesian computation (ABC) approaches will be used to estimate parameters of mathematical models and
compare our model with alternative ones. A parallel implementation is of our interests 
and performance data and inference outputs from ABC will be analysed then to evaluate the result.

@@

\begin{equation} \label{eq:1}
\begin{array}{l}
\frac{\mathrm{d} N}{\mathrm{d} t}=\lambda_N+\kappa_{N\beta}\beta-\mu_NN-\nu_{N\Phi}N\Phi\\\\
\frac{\mathrm{d} \Phi}{\mathrm{d} t}=\lambda_\Phi+\kappa_{\Phi\beta}\beta-\mu_\Phi\Phi\\\\
\frac{\mathrm{d} \beta}{\mathrm{d} t}=\frac{s_{\beta N}N}{1+i_{\beta\Phi}\Phi}-\mu_\beta\beta\\\\
\frac{\mathrm{d} \alpha}{\mathrm{d} t}=s_{\alpha\Phi}\Phi-\mu_\alpha\alpha
\end{array}
\end{equation}

A mathematical model will introduce parameters (e.g. $\kappa$, $\mu$ ,etc in (\ref{eq:1})), which are to be determined to 
give a specific model using observed data. A traditional method to find the best desired parameter sets 
is to apply an optimisation-based fitting, where the discrepancy of the model’s 
predictions and experimental data is minimised. Alternatively, by adopting Bayesian 
inference, we can focus on the experimental data and try to infer the parameter sets. 
Many approaches have been proposed and used in the area of systems biology\cite{ref:Stumpf}. Compared 
to optimisation-based fitting methods, Bayesian inference approaches can relieve the 
concerns of over-fitting and local optima. It focuses on posterior distribution which 
represents parameters’ ‘uncertainty’ given the observation data. As Bayes rule 
stated, the posterior distribution $p(\theta|D)$ is proportional to product of likelihood and 
prior
\begin{equation} \label{eq:2}
p(\theta|D)\propto l(\theta|D)p(\theta)
\end{equation}
where $\theta$ denotes parameter vector, $l$ is the likelihood function and $D$ is the observed data.

@@

\begin{enumerate}
	\item pyABC
	\begin{itemize}
		\item It has built-in support for multiple types of parallelisation, e.g., 
		multicore processing and distributed cluster

		\item Stored runs can be resumed

		\item Several built-in visualisation for the results are provided
	\end{itemize}

	\item astroABC
	\begin{itemize}
		\item It supports parallel sampling using MPI or multiprocessing

		\item Every run of the sampling can be paused and resumed later
	\end{itemize}

	\item ELFI
	\begin{itemize}
		\item It allows user to connect to \texttt{ipyparallel} for more parallelisation options

		\item It uses directed acyclic graph as input model; a conversion is needed before implementation
	\end{itemize}

	\item ABC-SysBio
	\begin{itemize}
		\item It has experimental CUDA accelerated parallelisation support
		\item It is specially designed for systems biology applications

	\item EasyABC
	\begin{itemize}
		\item It supports multicore parallelisation

		\item Several different sequential algorithms are offered
	\end{itemize}

	\item smfsb
	\begin{itemize}
		\item It provides simple and basic output of parameter estimations
	\end{itemize}
